Hierarchy Builder

### 2020-06-19 coqws/easychair.tex
Improving the presentation.

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}
\usepackage{wrapfig}
\usepackage[font=small,labelfont=bf]{caption}

\usepackage{doc}
\usepackage{minted} % pip3 install pygments, pdflatex -shell-escape hb
\newminted[elpicode]{elpi.py:ElpiLexer -x}{linenos=true,fontsize=\footnotesize,escapeinside=\%\%}
\newmintinline[elpi]{elpi.py:ElpiLexer -x}{fontsize=\small}
\newminted[coqcode]{elpi.py:CoqElpiLexer -x}{firstnumber=last,linenos=true,fontsize=\footnotesize,mathescape,escapeinside=\#\#}
\newmintinline[coq]{elpi.py:CoqElpiLexer -x}{fontsize=\small}
\definecolor{dkgreen}{rgb}{0, 0.502, 0}
\bibliographystyle{plainurl}% the mandatory bibstyle

@@

There is no questioning the importance of abstract algebra in the mathematical
vernacular. And it is with no surprise that libraries of machine checked code
are, nowadays, organized around hierarchies of structures. Unfortunately the
language of Type Theory and the features provided by the Coq system make the
construction of a hierarchy hard even for expert users. The difficulty
begins with the non-orthogonal choices of the encoding style, from bundled to
unbundled, and inference mechanism, from Type Classes to Canonical Structures.
To this, one may add the concerns about performance and about
the usability, by a non expert, of the final hierarchy.

In this talk we give an extensive demo of the Hierarchy Builder (\HB{})
extension for the Coq system version 8.10 and above.
\HB{} gives the library designer a language to describe the building blocks
of algebraic structures and to assemble them into a hierarchy. Similarly it
provides the final user linguistic constructs to build instances (examples) of
structures and to teach the elaborator of Coq how to take advantage of this
knowledge during type inference. Finally \HB{} lets the library designer
improve the usability of his library by providing alternative interfaces
to the primitive ones, a feature that can also be used to accommodate changes
to the hierarchy without breaking user code.

The details of \HB{}, including its compilation to the (Flat) Packed Classes
discipline and its implementation in the Elpi extension language for Coq,
are given in~\cite{cohen:hal-02478907}.
In this talk we focus on how to use \HB{} in a Coq development,
% and we plan to not detail how it works behind the scenes, 
unless questions on its internals arise from the audience.\\
\HB{} is available for download at \url{https://github.com/math-comp/hierarchy-builder}.

@@

\begin{coqcode}
HB.mixin Record CMonoid_of_Type A := { (* The set of axioms making A a commutative monoid. *)
  zero  : A;
  add   : A -> A -> A;
  addrA : associative add;  (* `add` is associative  *)
  addrC : commutative add;  (* `add` is commutative  *)
  add0r : left_id zero add; (* `zero` is the neutral *)
}.
HB.structure Definition CMonoid := { A of CMonoid_of_Type A }. (* The structure thereof. *)
Notation "0" := zero.
Infix    "+" := add.

(* The type of the operations and axioms depend on a CMonoid.type structure. *)
Check addrC. (* ?M : CMonoid.type |- forall x y : ?M, x + y = y + x *)

HB.mixin Record AbelianGrp_of_CMonoid A of CMonoid A := {
  (* We can write `add` here since A is a  CMonoid   *)
  opp   : A -> A;
  addNr : left_inverse zero opp add; (* `opp` is the additive inverse *)
}.
\end{coqcode}

\newpage

\begin{wrapfigure}[0]{r}{.35\textwidth}
  \vspace{1em}
  \includegraphics[width=.35\textwidth]{diagram.pdf}
  \vspace{-1.5em}
  \caption{The hierarchy}
  \label{fig:hierarchy}
\end{wrapfigure}
\
\begin{coqcode}
HB.structure Definition AbelianGrp := { A of AbelianGrp_of_CMonoid A & }.
Notation "- x"   := (opp x).
Notation "x - y" := (add x (opp y)).

(* A statement in the signature of an Abelian group G. *)
Check forall G : AbelianGrp.type, forall x : G, x - x = 0.

HB.mixin Record SemiRing_of_CMonoid A of CMonoid A := {
  one    : A;
  mul    : A -> A -> A;
  mulrA  : associative mul;  (* `mul` is associative   *)
  mul1r  : left_id one mul;  (* `one` is left neutral  *)
  mulr1  : right_id one mul; (* `one` is right neutral *)
  mulrDl : left_distributive mul add;  (* `mul` distributes over *)
  mulrDr : right_distributive mul add; (*   `add` on both sides  *)
  mul0r  : left_zero zero mul;  (* `zero` is absorbing `mul`     *)
  mulr0  : right_zero zero mul; (*   on both sides               *)
}.
HB.structure Definition SemiRing := { A of SemiRing_of_CMonoid A & }.
Notation "1"  := one.
Infix    "*"  := mul.

(* A statement in the signature of a Semiring S.  *)
Check forall S : SemiRing.type, forall x : S, x * 1 + 0 = x.

(* We need no additional mixin to declare the Ring structure. *)
HB.structure Definition Ring := { A of SemiRing_of_CMonoid A & AbelianGrp_of_CMonoid A }.

(* A statement in the signature of a Ring R.  *)
Check forall R : Ring.type, forall x : R, x * - (1) = - x.

(* We equip Z with CMonoid, AbelianGrp and SemiRing structures; HB equips it with Ring. *)
Definition Z_CMonoid    := CMonoid_of_Type.Build Z 0%Z Z.add Z.add_assoc Z.add_comm Z.add_0_l.
HB.instance Z Z_CMonoid.
Definition Z_AbelianGrp := AbelianGrp_of_CMonoid.Build Z Z.opp Z.add_opp_diag_l.
HB.instance Z Z_AbelianGrp.
Definition Z_SemiRing   := SemiRing_of_CMonoid.Build Z 1%Z Z.mul Z.mul_assoc Z.mul_1_l ...
HB.instance Z Z_SemiRing.

(* A statement in the signature of the Z ring *)
Check forall x : Z, x * - (1 + x) = 0 + 1.
\end{coqcode}

For space constraints we omit the declaration
of the green arrows in Figure~\ref{fig:hierarchy} which are alternative interfaces
called factories. They provide specialized interfaces to
build structures from others.
Factories are used to accommodate the insertion of
intermediate structures in a hierarchy without breaking user code, as well
as to implement shortcuts: e.g. the factory to build a \coq{Ring} from a \coq{Type}
(rightmost green arrow)
can omit the commutativity axiom of the addition, since it is implied
by the other axioms~\cite{nearrings}.
\bibliography{biblio}

>>>>

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}
\usepackage{amsmath, amssymb}

\usepackage{wrapfig}
\usepackage[font=small,labelfont=bf]{caption}

\usepackage{doc}
\usepackage{minted} % pip3 install pygments, pdflatex -shell-escape hb
\newminted[elpicode]{elpi.py:ElpiLexer -x}{linenos=true,fontsize=\footnotesize,escapeinside=\%\%}
\newmintinline[elpi]{elpi.py:ElpiLexer -x}{fontsize=\small}
\newminted[coqcode]{elpi.py:CoqElpiLexer -x}{firstnumber=last,linenos=false,fontsize=\footnotesize,mathescape,escapeinside=\#\#}
\newmintinline[coq]{elpi.py:CoqElpiLexer -x}{fontsize=\small}
\definecolor{dkgreen}{rgb}{0, 0.502, 0}
\bibliographystyle{plainurl}% the mandatory bibstyle

@@

Abstract algebra is extremely important in the mathematical vernacular
and libraries of machine checked code are, nowadays, organized around
hierarchies of structures. Unfortunately the language of Type Theory
and the features provided by the Coq system make the construction of a
hierarchy hard even for expert users. The difficulty begins with the
non-orthogonal choices, between storing information as record fields
or parameters, and between using type classes and canonical structures
for inference. To this, one may add the concerns about performance and
about the usability, by a non expert, of the final hierarchy.

\HB{} gives the library designer a language to describe the building
blocks of algebraic structures and to assemble them into a
hierarchy. Similarly it provides the final user linguistic constructs
to build instances (examples) of structures and to teach the
elaborator of Coq how to take advantage of this knowledge during type
inference. Finally \HB{} lets the library designer improve the
usability of his library by providing alternative interfaces to the
primitive ones, a feature that can also be used to accommodate changes
to the hierarchy without breaking user code. The details of \HB{},
including its compilation to a variant of the Packed Classes
discipline and its implementation in the Elpi extension language for
Coq, are given in~\cite{cohen:hal-02478907}.

In this talk we focus on how to use \HB{} in a Coq development without
any prior knowledge, walking slowly through the example below,
describing the bricks of the language from mixins -- i.e. bare bone building
block of the hierarchy, it packs operations and axioms -- to structures.

\HB{} is available for download at \url{https://github.com/math-comp/hierarchy-builder}.

@@

We first introduce the base structure of our example,
\verb-CMonoid-, which is defined in two phases: first its contents
using \verb-HB.mixin- and then the structure itself is declared using
\verb-HB.structure.-

\begin{coqcode}
HB.mixin Record CMonoid_of_Type A := { (* The set of axioms making A a commutative monoid. *)
  zero  : A; add   : A -> A -> A;
  addrA : associative add;  (* `add` is associative  *)
  addrC : commutative add;  (* `add` is commutative  *)
  add0r : left_id zero add; (* `zero` is a neutral element *)
}.
HB.structure Definition CMonoid := { A of CMonoid_of_Type A }. (* The structure thereof. *)
Notation "0" := zero.
Infix    "+" := add.

(* The type of the operations and axioms depend on a CMonoid.type structure. *)
Check addrC. (* ?M : CMonoid.type |- commutative (@add ?M) *)
\end{coqcode}

We extend this hierarchy with abelian group, semi ring and ring
structures, which happen to form a diamond (cf
Figure~\ref{fig:hierarchy}), using only the bare minimum information:
we use the same pattern as before to create the abelian group and
semi-ring structures, then a ring being no more than the combination
of the former structures, we do not even need to provide a mixin.

\begin{wrapfigure}[0]{r}{.35\textwidth}
  \vspace{-27em}
  \includegraphics[width=.35\textwidth]{diagram.pdf}
  \vspace{-1.5em}
  \caption{The hierarchy}
  \label{fig:hierarchy}
\end{wrapfigure}

\begin{coqcode}
HB.mixin Record AbelianGrp_of_CMonoid A of CMonoid A := {
  opp   : A -> A;
  (* We can write `add` here since A is a  CMonoid   *)
  addNr : left_inverse zero opp add; (* `opp` is the additive inverse *)
}.
HB.structure Definition AbelianGrp := { A of AbelianGrp_of_CMonoid A }.
Notation "- x"   := (opp x).
Notation "x - y" := (add x (opp y)).

HB.mixin Record SemiRing_of_CMonoid A of CMonoid A := {
  one    : A;
  mul    : A -> A -> A;
  mulrA  : associative mul;  (* `mul` is associative   *)
  mul1r  : left_id one mul;  (* `one` is left neutral  *)
  mulr1  : right_id one mul; (* `one` is right neutral *)
  mulrDl : left_distributive mul add;  (* `mul` distributes over *)
  mulrDr : right_distributive mul add; (*   `add` on both sides  *)
  mul0r  : left_zero zero mul;  (* `zero` is absorbing `mul`     *)
  mulr0  : right_zero zero mul; (*   on both sides               *)
}.
HB.structure Definition SemiRing := { A of SemiRing_of_CMonoid A }.
Notation "1"  := one.
Infix    "*"  := mul.

(* We need no additional mixin to declare the Ring structure. *)
HB.structure Definition Ring := { A of SemiRing_of_CMonoid A & AbelianGrp_of_CMonoid A }.
\end{coqcode}

Remark we can write abstract statements combining successive signatures:
\begin{coqcode}
(* An example statement in the signature of an Abelian group G, combining 0 and -. *)
Check forall G : AbelianGrp.type, forall x : G, x - x = 0.
(* An example statement in the signature of a Semiring S, combining 0, +, and *.  *)
Check forall S : SemiRing.type, forall x : S, x * 1 + 0 = x.
(* An example statement in the signature of a Ring R, combining -, 1 and *.  *)
Check forall R : Ring.type, forall x y : R, x * - (1 * y) = - x * y.
\end{coqcode}

Now, we can equip $\mathbb{Z}$ with commutative monoid, abelian group and
semi ring instances, and \HB{} equips it automatically with the ring
one, because there is only one possible choice.

\begin{coqcode}
HB.instance Definition Z_CMonoid    := CMonoid_of_Type.Build Z 0%Z Z.add
  Z.add_assoc Z.add_comm Z.add_0_l.
HB.instance Definition Z_AbelianGrp := AbelianGrp_of_CMonoid.Build Z Z.opp Z.add_opp_diag_l.
HB.instance Definition Z_SemiRing   := SemiRing_of_CMonoid.Build Z 1%Z Z.mul
  Z.mul_assoc Z.mul_1_l Z.mul_1_r Z.mul_add_distr_r Z.mul_add_distr_l Z.mul_0_l Z.mul_0_r.

(* An example statement in the signature of the Z ring, combining Z, 0, +, -, 1 and * *)
Check forall x : Z, x * - (1 + x) = 0 + 1.
\end{coqcode}

For space constraints we omit the declaration of the dashed arrows in
Figure~\ref{fig:hierarchy}, which are Factories, i.e. alternative
ways of building the same structures, but from different inputs.

Factories are used to accommodate the insertion of intermediate
structures in a hierarchy without breaking user code, as well as to
implement shortcuts: e.g. the factory to build a \coq{Ring} from a
\coq{Type} (rightmost dashed arrow) can omit the commutativity axiom
of the addition, since it is implied by the other axioms.

\bibliography{biblio}
